---
title: "Exploratory Data Analysis"
class-date: "Sep 16, 2026"
---

**Class date:** {{< meta class-date >}}

## Before We Get Started

<iframe src="slides/W03_EDA.pdf" width="100%" height="750px" style="border: 1px solid #ddd; border-radius: 8px;">

</iframe>

## Learning objectives

By the end of this session, you should be able to:

- distinguish numeric and categorical variables and identify the unit of observation in a dataset;
- perform a basic data audit before analysis;
- use descriptive statistics and simple visual summaries to understand a dataset;
- identify common data-quality problems, including missing values, duplicates, invalid codes, inconsistent categories, and unusual values;
- make defensible preprocessing decisions rather than automatically deleting problematic observations;
- transform dates, text, and categorical variables into analysis-ready formats;
- reshape and export a cleaned dataset that can be used for visualization in the next class.

::: callout-important
## Main idea

**Do not start by making a graph. Start by asking whether the data are trustworthy and whether each variable means what you think it means.**

A useful workflow is:

**Question → Audit → Clean → Explore → Reshape → Visualize → Communicate**
:::

# 1. Before EDA: understand what one row means

Before calculating statistics or drawing plots, identify the **unit of observation**.

Ask:

1.  What does one row represent?
2.  What does one column represent?
3.  Which columns are identifiers or metadata?
4.  Which variables are outcomes, predictors, or descriptive attributes?
5.  Is the dataset already aggregated, or does each row represent an individual event/person/case?

This matters because the same variable can mean different things at different levels of analysis.

For example:

- one row = one person;
- one row = one survey response;
- one row = one social-media post;
- one row = one traffic collision;
- one row = one country-year.

# 2. Variable types

## Numeric variables

Numeric variables have values that describe a measurable quantity: **how many** or **how much**.

- **Continuous numeric variable:** observations can take values on a continuous scale, such as temperature, income, or distance.
- **Discrete numeric variable:** observations represent counts, such as number of clicks, number of injuries, or number of purchases.

## Categorical variables

Categorical variables describe a characteristic or group membership: **what type** or **which category**.

- **Nominal variable:** categories have no meaningful ordering, such as borough or vehicle type.
- **Ordinal variable:** categories can be logically ordered, such as satisfaction level or education level.

::: callout-tip
## Numbers are not always numeric variables

A variable can be stored as numbers but still be categorical.

For example, if `1 = Seoul`, `2 = Busan`, and `3 = Incheon`, the numbers are **codes**, not quantities. Computing their mean would not be meaningful.
:::

# 3. What is exploratory data analysis (EDA)?

EDA is the process of learning what a dataset contains before making strong conclusions from it.

EDA usually combines:

1.  **Descriptive statistics** — numerical summaries of distributions and relationships.
2.  **Graphical summaries** — visual displays that reveal patterns that a few summary statistics may hide.
3.  **Data-quality checks** — checks for missing, duplicated, invalid, inconsistent, or unusual values.

EDA and preprocessing are therefore closely connected. We often discover cleaning problems **while exploring the data**.

# 4. A practical data-audit checklist

Before analysis, ask the following questions.

1.  **Rows:** How many observations are there?
2.  **Columns:** How many variables are there?
3.  **Variable roles:** Which columns are identifiers, attributes, outcomes, or metadata?
4.  **Data types:** Which variables are numeric, categorical, text, dates, or times?
5.  **Ranges:** Are minimum and maximum values plausible?
6.  **Categories:** Are category labels consistent and documented?
7.  **Missing values:** Where are data missing, and how much is missing?
8.  **Duplicates:** Are rows or identifiers unexpectedly duplicated?
9.  **Invalid values:** Are there impossible or undefined codes?
10. **Outliers:** Are extreme values errors, or are they valid but rare cases?

::: callout-note
## A useful principle

A preprocessing decision should always answer two questions:

1.  **What is wrong or inconvenient about the current value?**
2.  **Why is the chosen correction appropriate for the analysis?**

Never delete or replace observations only because a function makes it easy to do so.
:::

# 5. Case study: NYC Motor Vehicle Collisions

For the preprocessing exercise, we will use an English-language public dataset from **NYC Open Data**.

**Dataset:** Motor Vehicle Collisions – Crashes\
**Provider:** New York City Police Department (NYPD)\
**Dataset page:** <https://data.cityofnewyork.us/d/h9gi-nx95>

Each row represents **one motor vehicle collision**.

We will use a fixed historical period so that everyone works with roughly the same data.

## 5.1 Load packages

```{r}
library(tidyverse) #install.packages("tidyverse")
library(lubridate) #install.packages("lubridate")
```

## 5.2 Download a reproducible sample

We will request up to 5,000 collisions from January–March 2025. The code saves a local copy after the first download, so later renders do not depend on the internet connection.

```{r}
dir.create("data", showWarnings = FALSE)

raw_file <- "data/nyc_collisions_2025_q1.csv"

if (file.exists(raw_file)) {
  collisions_raw <- read_csv(raw_file, show_col_types = FALSE)
} else {
  raw_url <- paste0(
    "https://data.cityofnewyork.us/resource/h9gi-nx95.csv?",
    "$limit=5000&",
    "$where=crash_date between '2025-01-01T00:00:00.000' ",
    "and '2025-03-31T23:59:59.999'&",
    "$order=crash_date DESC"
  )

  collisions_raw <- read_csv(URLencode(raw_url), show_col_types = FALSE)
  write_csv(collisions_raw, raw_file)
}
```

## 5.3 First look

```{r}
dim(collisions_raw)
names(collisions_raw)
glimpse(collisions_raw)
head(collisions_raw)
```

Do not immediately scroll through thousands of rows. Start with structure.

```{r}
collisions_raw %>%
  select(
    collision_id, crash_date, crash_time, borough, zip_code,
    latitude, longitude, on_street_name,
    number_of_persons_injured, number_of_persons_killed,
    contributing_factor_vehicle_1, vehicle_type_code1
  ) %>%
  glimpse()
```

## 5.4 Keep variables we need for class

A large dataset often contains columns that are not relevant to the current question. Selecting a smaller working dataset makes later steps easier to understand.

```{r}
collisions <- collisions_raw %>%
  select(
    collision_id,
    crash_date,
    crash_time,
    borough,
    zip_code,
    latitude,
    longitude,
    on_street_name,
    number_of_persons_injured,
    number_of_persons_killed,
    contributing_factor_vehicle_1,
    vehicle_type_code1
  ) %>%
  rename(
    injured = number_of_persons_injured,
    killed = number_of_persons_killed,
    contributing_factor = contributing_factor_vehicle_1,
    vehicle_type = vehicle_type_code1
  )

glimpse(collisions)
```

# 6. Step 1 — Check dimensions, names, and data types

Basic questions:

- How many rows do we have?
- How many columns?
- Are variable names understandable?
- Did R read each variable using the expected data type?

```{r}
nrow(collisions)
ncol(collisions)
names(collisions)
glimpse(collisions)
```

A quick compact summary:

```{r}
collisions %>%
  summarise(
    rows = n(),
    unique_ids = n_distinct(collision_id),
    earliest_date = min(crash_date, na.rm = TRUE),
    latest_date = max(crash_date, na.rm = TRUE)
  )
```

::: callout-warning
`crash_date` may still be stored as text at this stage. We will explicitly convert it later instead of assuming that automatic type detection was correct.
:::

# 7. Step 2 — Missing values

Missingness is one of the most common problems in real project data.

## 7.1 Count missing values by variable

```{r}
missing_summary <- collisions %>%
  summarise(across(everything(), ~ sum(is.na(.)))) %>%
  pivot_longer(
    cols = everything(),
    names_to = "variable",
    values_to = "n_missing"
  ) %>%
  mutate(
    pct_missing = round(100 * n_missing / nrow(collisions), 1)
  ) %>%
  arrange(desc(pct_missing))

missing_summary
```

A quick visualization of missingness:

```{r}
ggplot(missing_summary,
       aes(x = reorder(variable, pct_missing), y = pct_missing)) +
  geom_col() +
  coord_flip() +
  labs(
    x = NULL,
    y = "Missing (%)",
    title = "Missing values by variable"
  ) +
  theme_minimal()
```

## 7.2 Missing values are not all the same

Suppose `borough` is missing.

Possible interpretations include:

- the collision occurred outside a coded borough;
- the location was unavailable;
- the field was not entered;
- the borough could possibly be recovered from another location variable.

Therefore, this is usually a poor default:

```{r, eval=FALSE}
# Do NOT automatically do this to the entire dataset.
collisions <- drop_na(collisions)
```

Instead, remove missing values **only when a specific analysis requires that variable**.

```{r}
borough_data <- collisions %>%
  filter(!is.na(borough))
```

The original dataset remains unchanged.

## 7.3 Explicit NA vs coded missing values

Some datasets contain values such as:

- `Unknown`
- `Not available`
- `N/A`
- `999`
- `-99`
- `Unspecified`

These may represent missingness even though R does not treat them as `NA`.

Check a categorical variable:

```{r}
collisions %>%
  count(contributing_factor, sort = TRUE) %>%
  slice_head(n = 15)
```

For this dataset, `Unspecified` is often more useful as an unknown/missing category than as a substantive contributing factor.

```{r}
collisions <- collisions %>%
  mutate(
    contributing_factor = na_if(contributing_factor, "Unspecified")
  )
```

Check again:

```{r}
collisions %>%
  summarise(
    missing_factor = sum(is.na(contributing_factor)),
    pct_missing_factor = round(mean(is.na(contributing_factor)) * 100, 1)
  )
```

# 8. Step 3 — Duplicates and identifiers

Duplicates can occur at two levels:

1.  **Exact duplicate rows** — every value is identical.
2.  **Duplicated identifiers** — the same ID occurs in more than one row.

These are not the same problem.

## 8.1 Exact duplicate rows

```{r}
sum(duplicated(collisions))
```

If exact duplicates exist, inspect them before deleting:

```{r}
collisions %>%
  filter(duplicated(.) | duplicated(., fromLast = TRUE)) %>%
  head()
```

Only after verifying that they are accidental duplicates would we use:

```{r, eval=FALSE}
collisions <- collisions %>% distinct()
```

## 8.2 Duplicated IDs

`collision_id` is intended to identify a collision.

```{r}
collisions %>%
  summarise(
    n_rows = n(),
    n_unique_ids = n_distinct(collision_id),
    duplicated_ids = n() - n_distinct(collision_id)
  )
```

If duplicated IDs are found:

```{r}
collisions %>%
  count(collision_id, sort = TRUE) %>%
  filter(n > 1)
```

::: callout-important
## Do not automatically keep the first duplicate

`distinct(collision_id, .keep_all = TRUE)` is convenient, but it silently discards information. First determine **why** the duplicate exists.
:::

# 9. Step 4 — Convert dates and times

Dates are often imported as character strings. Convert them explicitly.

```{r}
collisions <- collisions %>%
  mutate(
    crash_date = ymd(crash_date),
    crash_time_parsed = hm(crash_time),
    crash_hour = hour(crash_time_parsed),
    crash_weekday = wday(
      crash_date,
      label = TRUE,
      abbr = FALSE,
      week_start = 1
    )
  )
```

Check the result:

```{r}
collisions %>%
  select(crash_date, crash_time, crash_hour, crash_weekday) %>%
  head()
```

Why bother?

Once dates and times are properly represented, we can easily ask questions such as:

- Are collisions more frequent on weekends?
- At what hour are collisions most common?
- Are severe collisions concentrated at particular times?

These become visualization questions for the next class.

# 10. Step 5 — Inspect ranges and impossible values

## 10.1 Numeric ranges

```{r}
collisions %>%
  summarise(
    min_injured = min(injured, na.rm = TRUE),
    max_injured = max(injured, na.rm = TRUE),
    min_killed = min(killed, na.rm = TRUE),
    max_killed = max(killed, na.rm = TRUE),
    min_latitude = min(latitude, na.rm = TRUE),
    max_latitude = max(latitude, na.rm = TRUE),
    min_longitude = min(longitude, na.rm = TRUE),
    max_longitude = max(longitude, na.rm = TRUE)
  )
```

## 10.2 Rule-based checks

Some impossible values can be identified from the variable definition itself.

For example, counts should not be negative.

```{r}
collisions %>%
  filter(injured < 0 | killed < 0)
```

Coordinates equal to `(0, 0)` are not located in New York City, so they should be treated as invalid coordinates rather than real locations.

```{r}
collisions %>%
  summarise(
    zero_latitude = sum(latitude == 0, na.rm = TRUE),
    zero_longitude = sum(longitude == 0, na.rm = TRUE)
  )
```

Convert those placeholder coordinates to missing values:

```{r}
collisions <- collisions %>%
  mutate(
    latitude = na_if(latitude, 0),
    longitude = na_if(longitude, 0)
  )
```

# 11. Step 6 — Outliers: inspect first, decide later

An **outlier** is an unusual observation. It is not automatically an error.

Possible reasons for an extreme value:

- data-entry error;
- measurement error;
- coding error;
- rare but genuine event;
- observation from a different population;
- unit mismatch.

## 11.1 Summary statistics

```{r}
collisions %>%
  summarise(
    n = n(),
    mean_injured = mean(injured, na.rm = TRUE),
    median_injured = median(injured, na.rm = TRUE),
    sd_injured = sd(injured, na.rm = TRUE),
    q1 = quantile(injured, .25, na.rm = TRUE),
    q3 = quantile(injured, .75, na.rm = TRUE),
    max_injured = max(injured, na.rm = TRUE)
  )
```

## 11.2 IQR rule as a flagging tool

The IQR rule can flag observations for inspection.

```{r}
q1 <- quantile(collisions$injured, .25, na.rm = TRUE)
q3 <- quantile(collisions$injured, .75, na.rm = TRUE)
iqr_value <- IQR(collisions$injured, na.rm = TRUE)
upper_fence <- q3 + 1.5 * iqr_value

injury_outliers <- collisions %>%
  filter(injured > upper_fence) %>%
  arrange(desc(injured))

injury_outliers %>%
  select(
    collision_id, crash_date, borough,
    injured, killed, vehicle_type, contributing_factor
  ) %>%
  head(20)
```

::: callout-warning
## Outlier does not mean delete

For collision data, a crash involving many injured people could be a real bus or multi-vehicle crash. Automatically removing it could erase substantively important information.

Treat the IQR rule as an **inspection rule**, not a deletion rule.
:::

## 11.3 When might we transform an extreme value?

Possible strategies include:

- keep it as observed;
- verify it against the original source;
- exclude it with a documented reason;
- cap/winsorize it for a specific analysis;
- transform a strongly skewed variable (for example, log transformation);
- use a robust statistic such as the median.

The choice depends on the research question.

# 12. Step 7 — Clean text and categories

Text data often contain inconsistency in capitalization, spaces, spelling, or labels.

## 12.1 Remove extra whitespace

```{r}
collisions <- collisions %>%
  mutate(
    across(
      c(borough, on_street_name, contributing_factor, vehicle_type),
      ~ str_squish(.x)
    )
  )
```

## 12.2 Inspect category frequencies

```{r}
collisions %>%
  count(borough, sort = TRUE)
```

```{r}
collisions %>%
  count(vehicle_type, sort = TRUE) %>%
  slice_head(n = 30)
```

A real dataset may contain several labels that describe broadly similar vehicle types.

## 12.3 Collapse categories when substantively reasonable

The exact rules should depend on the analytical purpose. For a simple classroom example:

```{r}
collisions <- collisions %>%
  mutate(
    vehicle_group = case_when(
      is.na(vehicle_type) ~ NA_character_,
      str_detect(str_to_lower(vehicle_type), "motorcycle") ~ "Motorcycle",
      str_detect(str_to_lower(vehicle_type), "bike|bicycle") ~ "Bike",
      str_detect(str_to_lower(vehicle_type), "taxi") ~ "Taxi",
      str_detect(str_to_lower(vehicle_type), "sedan") ~ "Sedan",
      str_detect(
        str_to_lower(vehicle_type),
        "sport utility|station wagon|suv"
      ) ~ "SUV / Station Wagon",
      TRUE ~ "Other"
    )
  )
```

Check the recoded variable:

```{r}
collisions %>%
  count(vehicle_group, sort = TRUE)
```

::: callout-note
## Recode with a reason

Collapsing categories loses information. It can be useful when categories are too sparse or inconsistent, but always keep the raw variable as well.
:::

# 13. Step 8 — Handling missing data: practical options

There is no single correct rule for missing data. Common options include:

## Option A. Keep missing values

Often the safest default during early EDA.

```{r}
collisions_clean <- collisions
```

## Option B. Filter missing values for a specific analysis

For a borough comparison:

```{r}
borough_analysis <- collisions_clean %>%
  filter(!is.na(borough))
```

## Option C. Create an explicit "Unknown" category

Useful when missingness itself is meaningful for descriptive reporting.

```{r}
collisions_clean <- collisions_clean %>%
  mutate(
    borough_display = replace_na(borough, "Unknown")
  )
```

## Option D. Imputation

Imputation replaces missing values using an estimated value.

Examples include:

- mean or median imputation;
- group-specific imputation;
- regression/model-based imputation;
- multiple imputation.

For this introductory class, **do not impute automatically**. The method must be justified by the variable, the missing-data mechanism, and the purpose of the analysis.

# 14. Step 9 — Create useful derived variables

Preprocessing is not only about fixing errors. It also includes creating variables that better represent the analytical question.

## 14.1 Severe collision indicator

```{r}
collisions_clean <- collisions_clean %>%
  mutate(
    severe_collision = injured > 0 | killed > 0
  )
```

## 14.2 Time of day

```{r}
collisions_clean <- collisions_clean %>%
  mutate(
    time_of_day = case_when(
      crash_hour >= 5  & crash_hour < 12 ~ "Morning",
      crash_hour >= 12 & crash_hour < 17 ~ "Afternoon",
      crash_hour >= 17 & crash_hour < 21 ~ "Evening",
      !is.na(crash_hour) ~ "Night",
      TRUE ~ NA_character_
    ),
    time_of_day = factor(
      time_of_day,
      levels = c("Morning", "Afternoon", "Evening", "Night")
    )
  )
```

Check:

```{r}
collisions_clean %>%
  count(time_of_day)
```

# 15. Step 10 — Reshape data for analysis and visualization

Data are often easier to visualize when they are in **long format**.

Suppose we want to compare injuries and fatalities.

```{r}
severity_long <- collisions_clean %>%
  select(collision_id, crash_date, borough, injured, killed) %>%
  pivot_longer(
    cols = c(injured, killed),
    names_to = "outcome",
    values_to = "count"
  )

severity_long %>% head()
```

Wide format:

| collision_id | injured | killed |
|--------------|--------:|-------:|
| 1            |       2 |      0 |
| 2            |       0 |      1 |

Long format:

| collision_id | outcome | count |
|--------------|---------|------:|
| 1            | injured |     2 |
| 1            | killed  |     0 |
| 2            | injured |     0 |
| 2            | killed  |     1 |

Long format is especially useful with `ggplot2` because variable names can be mapped directly to aesthetics such as `x`, `colour`, `fill`, or `facet`.

# 16. Step 11 — Optional: one-hot encoding

One-hot encoding is mainly useful when categorical variables must be represented numerically for a statistical or machine-learning model.

For visualization, keeping a categorical variable as a factor is usually more useful.

```{r}
model_matrix <- model.matrix(
  ~ borough_display + time_of_day + vehicle_group - 1,
  data = collisions_clean
)

head(model_matrix)
```

::: callout-tip
## For this course

Do not one-hot encode categories just because you can. Use it when the next analytical method requires numeric design columns.
:::

# 17. Step 12 — Final quality check

After cleaning, repeat the audit.

```{r}
collisions_clean %>%
  summarise(
    rows = n(),
    unique_ids = n_distinct(collision_id),
    missing_borough = sum(is.na(borough)),
    missing_coordinates = sum(is.na(latitude) | is.na(longitude)),
    missing_vehicle_group = sum(is.na(vehicle_group)),
    min_injured = min(injured, na.rm = TRUE),
    max_injured = max(injured, na.rm = TRUE)
  )
```

Check the structure again:

```{r}
glimpse(collisions_clean)
```

# 18. Keep a preprocessing log

A good project should document what changed.

For example:

```{r}
preprocessing_log <- tribble(
  ~step, ~decision,
  "Missing factor codes", "Converted 'Unspecified' to NA",
  "Coordinates", "Converted zero latitude/longitude placeholders to NA",
  "Date/time", "Parsed crash date and created crash hour/weekday",
  "Vehicle categories", "Created broad vehicle_group while retaining raw vehicle_type",
  "Outliers", "Flagged using IQR rule; did not automatically delete",
  "Missing borough", "Retained in master data; filter only for borough-specific analyses"
)

preprocessing_log
```

This is useful for:

- reproducibility;
- team projects;
- explaining analytical decisions;
- debugging later;
- writing a methods section.

# 19. Save the cleaned dataset

```{r}
write_csv(
  collisions_clean,
  "data/nyc_collisions_2025_q1_clean.csv"
)
```

Next week, we can load the cleaned dataset directly:

```{r, eval=FALSE}
collisions <- read_csv(
  "data/nyc_collisions_2025_q1_clean.csv",
  show_col_types = FALSE
)
```

# 20. Quick EDA after preprocessing

Now that the data are more usable, descriptive summaries become easier to interpret.

## One numeric variable

```{r}
collisions_clean %>%
  summarise(
    mean = mean(injured, na.rm = TRUE),
    median = median(injured, na.rm = TRUE),
    sd = sd(injured, na.rm = TRUE),
    min = min(injured, na.rm = TRUE),
    max = max(injured, na.rm = TRUE)
  )
```

## One categorical variable

```{r}
collisions_clean %>%
  count(borough_display, sort = TRUE) %>%
  mutate(percent = round(100 * n / sum(n), 1))
```

## Two variables

```{r}
collisions_clean %>%
  group_by(borough_display) %>%
  summarise(
    collisions = n(),
    avg_injured = mean(injured, na.rm = TRUE),
    severe_rate = mean(severe_collision, na.rm = TRUE)
  ) %>%
  arrange(desc(collisions))
```

# 21. A small preview of visualization

We will cover chart selection and visual design in detail next week. For now, the goal is simply to verify that the cleaned data behave as expected.

## Collisions by borough

```{r}
collisions_clean %>%
  filter(!is.na(borough)) %>%
  ggplot(aes(x = borough)) +
  geom_bar() +
  labs(
    x = NULL,
    y = "Number of collisions",
    title = "Collisions by borough"
  ) +
  theme_minimal()
```

## Collisions by hour

```{r}
collisions_clean %>%
  filter(!is.na(crash_hour)) %>%
  count(crash_hour) %>%
  ggplot(aes(x = crash_hour, y = n)) +
  geom_line() +
  geom_point() +
  scale_x_continuous(breaks = 0:23) +
  labs(
    x = "Hour of day",
    y = "Number of collisions",
    title = "Collisions by time of day"
  ) +
  theme_minimal()
```

::: callout-note
## Stop here for today

The plots above are intentionally simple. Next week, we will ask a different question:

> **Given a question and variable types, which chart should we choose, and how should we design it?**

That is where chart choice, visual encodings, scales, labels, colour, faceting, and communication become the focus.
:::

# 22. In-class practice: audit a dataset

Work in pairs. Choose one dataset that your team may realistically use for the final project, or use the NYC collision data.

Complete the following checklist.

## A. Structure

- What is one row?
- How many rows and columns?
- Which variable is the identifier?
- Which variables are numeric, categorical, dates, or text?

## B. Missingness

- Which three variables have the most missing values?
- Are any missing values encoded using a special string or number?
- Would you keep, filter, replace, or investigate them?

## C. Duplicates

- Are there exact duplicate rows?
- Is the identifier unique?
- If not, what might explain the duplication?

## D. Validity

- Check minimum and maximum values for important numeric variables.
- Check category frequencies.
- Are there impossible or undefined values?

## E. Outliers

- Identify at least one extreme observation.
- Is it likely to be an error or a valid rare event?
- What evidence would you need before removing it?

## F. Preprocessing decision

Create a small table with three columns:

| Problem | Decision | Reason |
|----|----|----|
| Example: missing borough | Keep in master data | Other variables are still usable |
| Example: `Unspecified` factor | Convert to NA | Represents unknown factor |

# 23. Project-ready preprocessing checklist

Before making final project visualizations, check the following.

- [ ] I can explain what one row represents.
- [ ] I know which variables are numeric, categorical, dates, and identifiers.
- [ ] I checked missing values variable by variable.
- [ ] I checked exact duplicates and duplicated IDs.
- [ ] I checked category labels and undefined codes.
- [ ] I checked plausible ranges for important numeric variables.
- [ ] I inspected outliers before deciding whether to remove them.
- [ ] I kept raw variables when creating cleaned or collapsed versions.
- [ ] I documented every important preprocessing decision.
- [ ] I saved a cleaned dataset separately from the raw dataset.
- [ ] My cleaned data are ready for the visualization question I want to answer.

# 24. Key takeaways

1.  **EDA is not separate from cleaning.** Exploration often reveals data-quality problems.
2.  **Missing does not automatically mean delete.** The appropriate decision depends on the variable and analysis.
3.  **Duplicate does not automatically mean error.** First understand the unit of observation and identifier.
4.  **Outlier does not automatically mean remove.** Rare events can be substantively important.
5.  **Categories require interpretation.** Numeric codes, text labels, and `Unspecified` values need different treatment.
6.  **Preprocessing should be reproducible.** Keep the raw data, write code for every transformation, and maintain a decision log.
7.  **Cleaning should serve the question.** The goal is not a perfectly clean dataset; it is a dataset that is appropriate for the intended analysis.
