---
title: "Interactive Data Visualization II: Plotly"
class-date: "Oct 7, 2026"
toc: true
---

**Class date:** {{< meta class-date >}}

## 1. Setup

```{r}
#| label: setup
#| message: false
#| warning: false
 
# install.packages("plotly")
library(tidyverse)
library(plotly)
library(lubridate)
```

## 2. Building a Plotly Chart

```{r}
data(iris)
glimpse(iris)
```

### Step 1. Start with the data

```{r}
plot_ly(iris)
```

### Step 2. Map variables to the axes and choose a chart type

```{r}
plot_ly(
  iris,
  x = ~Sepal.Length,
  y = ~Petal.Length,
  type = "scatter"
)
```

### Step 3. Add a display mode

```{r}
plot_ly(
  iris,
  x = ~Sepal.Length,
  y = ~Petal.Length,
  type = "scatter",
  mode = "markers"
)
```

For `type = "scatter"`:

``` r
mode = "markers"        # points
mode = "lines"          # lines
mode = "lines+markers"  # lines and points
```

### Step 4. Map a grouping variable to color

```{r}
plot_ly(
  iris,
  x = ~Sepal.Length,
  y = ~Petal.Length,
  color = ~Species,
  type = "scatter",
  mode = "markers"
)
```

### Step 5. Add useful hover information

```{r}
plot_ly(
  iris,
  x = ~Sepal.Length,
  y = ~Petal.Length,
  color = ~Species,
  type = "scatter",
  mode = "markers",
  text = ~paste0(
    "Species: ", Species,
    "<br>Sepal length: ", Sepal.Length,
    "<br>Petal length: ", Petal.Length
  ),
  hoverinfo = "text"
)
```

### Step 6. Refine the appearance

```{r}
plot_ly(
  iris,
  x = ~Sepal.Length,
  y = ~Petal.Length,
  color = ~Species,
  type = "scatter",
  mode = "markers",
  text = ~paste0(
    "Species: ", Species,
    "<br>Sepal length: ", Sepal.Length,
    "<br>Petal length: ", Petal.Length
  ),
  hoverinfo = "text",
  marker = list(size = 9, opacity = .7)
  ) %>%
  layout(
    title = "Sepal Length and Petal Length",
    xaxis = list(title = "Sepal Length"),
    yaxis = list(title = "Petal Length")
  )
```

## 3. The Basic Structure of `plot_ly()`

``` r
plot_ly(
  data = ...,
  x = ~...,
  y = ~...,
  color = ~...,
  type = "...",
  mode = "..."
) %>%
  layout(...)
```

| Component           | Role                                           |
|---------------------|------------------------------------------------|
| `data`              | Which dataset?                                 |
| `x`, `y`            | Which variables?                               |
| `type`              | Which chart family?                            |
| `mode`              | How should observations be displayed?          |
| `color`, `size`     | How should additional variables be encoded?    |
| `text`, `hoverinfo` | What should appear on hover?                   |
| `layout()`          | Titles, axes, legend, and overall presentation |

### Common Plotly chart types

| `type`        | Chart                     |
|---------------|---------------------------|
| `"scatter"`   | Scatterplot or line chart |
| `"bar"`       | Bar chart                 |
| `"box"`       | Box plot                  |
| `"violin"`    | Violin plot               |
| `"histogram"` | Histogram                 |
| `"pie"`       | Pie or donut chart        |
| `"heatmap"`   | Heatmap                   |
| `"contour"`   | Contour plot              |
| `"surface"`   | 3D surface plot           |
| `"scatter3d"` | 3D scatterplot            |

## 4. Practice with `state.x77`

```{r}
states <- as.data.frame(state.x77) %>%
  rownames_to_column("State")

glimpse(states)

states <- states %>%
  rename(Life.Exp = `Life Exp`)
```

### Interactive scatterplot

```{r}
plot_ly(
  data = states,
  x = ~Income,
  y = ~Life.Exp,
  type = "scatter",
  mode = "markers",
  text = ~paste0(
    "State: ", State,
    "<br>Income: ", Income,
    "<br>Life expectancy: ", round(Life.Exp, 1)
  ),
  hoverinfo = "text",
  marker = list(size = 10, opacity = .7)
) %>%
  layout(
    title = "Income and Life Expectancy across U.S. States",
    xaxis = list(title = "Income"),
    yaxis = list(title = "Life Expectancy")
  )
```

### Add a third numeric variable using marker size

```{r}
plot_ly(
  data = states,
  x = ~Income,
  y = ~Life.Exp,
  size = ~Population,
  sizes = c(8, 30),
  type = "scatter",
  mode = "markers",
  text = ~paste0(
    "State: ", State,
    "<br>Income: ", Income,
    "<br>Life expectancy: ", round(Life.Exp, 1),
    "<br>Population: ", Population
  ),
  hoverinfo = "text"
) %>%
  layout(
    title = "Income, Life Expectancy, and Population",
    xaxis = list(title = "Income"),
    yaxis = list(title = "Life Expectancy")
  )
```

## 5. Practice with `airquality`

```{r}
data(airquality)
glimpse(airquality)

air_plot <- airquality %>%
  mutate(Month = factor(Month,
                        levels = 5:9,
                        labels = c("May", "June", "July", 
                                   "August", "September")))
head(air_plot$Month)
```

### Interactive Box Plot

```{r}
plot_ly(
  data = air_plot,
  x = ~Month,
  y = ~Temp,
  color = ~Month,
  type = "box",
  boxpoints = "outliers"
  ) %>%
  layout(title = "Temperature Distribution by Month",
         xaxis = list(title = ""),
         yaxis = list(title = "Temperature (°F)"),
         showlegend = FALSE)
```

## 5. Load the cleaned data (NYC Collision)

```{r}
df <- read_csv("data/nyc_collisions_clean.csv")
glimpse(df)
```

## Interactive Box Plot

Box plots can also be interactive.

Here, we use the built-in `airquality` dataset.

```{}
```

## 6. Interactive Bar Chart

```{r}
borough_counts <- df %>%
  filter(!is.na(borough), borough != "") %>%
  count(borough)
```

```{r}
plot_ly(
  data = borough_counts,
  x = ~borough,
  y = ~n,
  type = "bar",
  text = ~paste0(
    "Borough: ", borough,
    "<br>Collisions: ", scales::comma(n)
  ),
  hoverinfo = "text"
) %>%
  layout(
    title = "Motor Vehicle Collisions by Borough",
    xaxis = list(title = ""),
    yaxis = list(title = "Number of Collisions")
  )
```

> Does hovering add meaningful information here, or is the bar chart already easy to read?

------------------------------------------------------------------------

## 7. Interactive Heatmap

A heatmap is useful for comparing patterns across two categorical dimensions.

```{r}
heatmap_data <- df %>%
  filter(
    !is.na(borough),
    borough != "",
    !is.na(time_of_day)
  ) %>%
  count(borough, time_of_day)
```

```{r}
plot_ly(
  data = heatmap_data,
  x = ~borough,
  y = ~time_of_day,
  z = ~n,
  type = "heatmap",
  text = ~paste0(
    "Borough: ", borough,
    "<br>Time of day: ", time_of_day,
    "<br>Collisions: ", scales::comma(n)
  ),
  hoverinfo = "text"
) %>%
  layout(
    title = "Collision Frequency by Borough and Time of Day",
    xaxis = list(title = "Borough"),
    yaxis = list(title = "Time of Day")
  )
```

> Higher collision counts may partly reflect differences in population, traffic volume, or road exposure across boroughs.

------------------------------------------------------------------------

## 8. Interactive Scatterplot

First, summarize the data by day.

```{r}
daily_summary <- df %>%
  filter(
    !is.na(crash_date),
    !is.na(injured)
  ) %>%
  group_by(crash_date) %>%
  summarise(
    collisions = n(),
    total_injured = sum(injured, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(
    weekday = wday(
      crash_date,
      label = TRUE,
      abbr = FALSE,
      week_start = 1
    ),
    day_type = if_else(
      weekday %in% c("Saturday", "Sunday"),
      "Weekend",
      "Weekday"
    )
  )
```

```{r}
plot_ly(
  data = daily_summary,
  x = ~collisions,
  y = ~total_injured,
  color = ~day_type,
  type = "scatter",
  mode = "markers",
  text = ~paste0(
    "Date: ", crash_date,
    "<br>Collisions: ", collisions,
    "<br>Injured: ", total_injured,
    "<br>Day type: ", day_type
  ),
  hoverinfo = "text",
  marker = list(
    size = 8,
    opacity = 0.6
  )
) %>%
  layout(
    title = "Daily Collision Volume and Injuries",
    xaxis = list(title = "Number of Collisions"),
    yaxis = list(title = "Number of Persons Injured"),
    legend = list(title = list(text = "Day Type"))
  )
```

> Do days with more collisions tend to have more injured persons?

------------------------------------------------------------------------

## 9. Interactive Time Series

```{r}
daily_collisions <- df %>%
  filter(!is.na(crash_date)) %>%
  count(crash_date, name = "n")
```

```{r}
plot_ly(
  data = daily_collisions,
  x = ~crash_date,
  y = ~n,
  type = "scatter",
  mode = "lines",
  text = ~paste0(
    "Date: ", crash_date,
    "<br>Collisions: ", n
  ),
  hoverinfo = "text"
) %>%
  layout(
    title = "Daily Motor Vehicle Collisions",
    xaxis = list(title = ""),
    yaxis = list(title = "Number of Collisions")
  )
```

Try zooming into a shorter time period.

------------------------------------------------------------------------

## 10. Animated Bubble Chart

Animation can be useful when several dimensions change together over time.

```{r}
monthly_borough <- df %>%
  filter(
    !is.na(crash_date),
    !is.na(borough),
    borough != "",
    !is.na(injured)
  ) %>%
  mutate(
    crash_date = ymd(crash_date),
    month = floor_date(crash_date, "month")
  ) %>%
  group_by(month, borough) %>%
  summarise(
    collisions = n(),
    total_injured = sum(injured, na.rm = TRUE),
    severe_collisions = sum(severe_collision, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(
    injured_per_100 = total_injured / collisions * 100,
    month_label = format(month, "%Y-%m")
  )
```

```{r}
plot_ly(
  data = monthly_borough,
  x = ~collisions,
  y = ~injured_per_100,
  size = ~total_injured,
  sizes = c(20, 80),
  color = ~borough,
  frame = ~month_label,
  text = ~paste0(
    "Borough: ", borough,
    "<br>Month: ", month_label,
    "<br>Collisions: ", scales::comma(collisions),
    "<br>Injured per 100 collisions: ", round(injured_per_100, 1),
    "<br>Total injured: ", scales::comma(total_injured)
  ),
  hoverinfo = "text",
  type = "scatter",
  mode = "markers",
  marker = list(
    opacity = 0.75,
    line = list(width = 1, color = "white")
  )
) %>%
  layout(
    title = "How Collision Patterns Change over Time",
    xaxis = list(
      title = "Number of Collisions",
      type = "log"
    ),
    yaxis = list(
      title = "Persons Injured per 100 Collisions",
      range = c(0, 40)
    ),
    legend = list(
      title = list(text = "Borough")
    )
  ) %>%
  animation_opts(
    frame = 200,
    transition = 100,
    redraw = FALSE
  ) %>%
  animation_slider(
    currentvalue = list(
      prefix = "Month: "
    )
  )
```

This figure encodes several dimensions at once:

- x position → collision volume
- y position → injury rate
- color → borough
- bubble size → total injured
- animation → time

------------------------------------------------------------------------

## 11. Static or Interactive?

| Static chart                 | Interactive chart                    |
|------------------------------|--------------------------------------|
| Fast to read                 | Allows exploration                   |
| Easy to print                | Provides details on demand           |
| Good for one clear message   | Useful for many observations         |
| Useful in reports and papers | Useful in dashboards and web reports |
| Author controls the view     | User can change the view             |

> Use interactivity when users benefit from **hovering, exploring, filtering, zooming, or requesting details**.

Do not add interaction simply because it is available.

------------------------------------------------------------------------

## 12. Mini Practice

Choose **one** of the following.

### Option A: `iris`

Create an interactive plot that examines the relationship between two flower measurements.

Requirements:

- choose two numeric variables;
- use `Species` as color;
- add useful hover information;
- add a title and axis labels.

### Option B: `state.x77`

Create an interactive plot using two numeric variables of your choice.

Requirements:

- identify each state in the hover text;
- use an appropriate chart type;
- add a title and axis labels.

### Option C: NYC collisions

Create one interactive visualization that answers a question about collision patterns.

Possible questions:

- How do collision counts differ across boroughs?
- How does collision frequency change over time?
- Are days with more collisions associated with more injuries?
- Do collision patterns differ between weekdays and weekends?

------------------------------------------------------------------------

## 13. Wrap-up

Today we focused on direct Plotly syntax:

``` r
plot_ly(
  data = ...,
  x = ~...,
  y = ~...,
  type = ...
)
```

Common additions include:

``` r
color = ~group
text = ~...
hoverinfo = "text"
mode = "markers"
```

> **The main lesson is not the syntax.**\
> Interactivity should make the visualization easier to explore or understand.

Before adding interaction, ask:

1.  What question am I answering?
2.  What chart type fits the variables?
3.  What additional information should appear on demand?
4.  Does interaction make the chart easier to understand?

> **Interactivity is not the goal; better understanding is.**
